---
title: "Change over time of fish community compositon and abundance at each site and park"
author: "MPeipoch"
date: "2023-12-03"
output: pdf_document
---

Necessary packages for the script
```{r, echo=FALSE}
# required packages
req_packs <- c("ggplot2", "dplyr", "vegan", "tidyr","stringr","openxlsx","sf","raster","trend","lubridate","FSA","cowplot")


########### if installed, then install and load ###################################
new_packs <- req_packs[!(req_packs %in% installed.packages()[,"Package"])]
if (length(new_packs) > 0) {
  install.packages(new_packs,dependencies=T)
}

lapply(req_packs, function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    stop(paste("Package", pkg, "could not be loaded. Fix it!"))
  }
  library(pkg, character.only = TRUE)
})




```



TEMPORAL CHANGE IN WATER QUALITY/HABITAT METRICS PER SITE & YEAR
This section uses the  the lastest '20231128_wqp_wqx_bss_wq_npsncrn.csv' dataset from Charlie to calculate temporal trends in habitat and water quality per sitre that we can relate to fish change over time. 
```{r}

###Import data file 

setwd("/Users/mpeipoch/Library/CloudStorage/GoogleDrive-mpeipoch@stroudcenter.org/My Drive/NPS_FishDataAnalysis")
wdata = read.csv("20231128_wqp_wqx_bss_wq_npsncrn.csv")


###Create a vector with the unique sites in USNPS NCRN Biological stream survey

sites = unique(wdata$MonitoringLocationIdentifier)



###Create lists of length=sites in which all available waterchem/habitat data and summary metrics per site is stored in a unique format and file

NCRNAnnualWatChem = list() # Results output


for (site in sites) {
  
  # Subset data for each site
  temp_data = wdata %>% 
    dplyr::filter(MonitoringLocationIdentifier == site)

    
        temp_data_subset = temp_data %>% #select variable per site and method 
          dplyr::filter(CharacteristicName %in% c("Salinity","Specific conductance","Temperature, water"))
          #add here the parameters of interest

 
out_data = temp_data_subset %>%
  mutate(samplingYear =  as.factor(year(temp_data_subset$ActivityStartDate))) %>%
    group_by(samplingYear,CharacteristicName) %>%
    summarise(AnnualMedian = median(as.numeric(ResultMeasureValue),na.rm=T),
              AnnualMean = mean(as.numeric(ResultMeasureValue),na.rm=T),
              AnnualCount = n(),
              AnnualSD = sd(as.numeric(ResultMeasureValue)),
              Annuallower_ci = AnnualMean - qt(0.975, df = n() - 1) * (AnnualSD / sqrt(n())),
              Annualupper_ci = AnnualMean + qt(0.975, df = n() - 1) * (AnnualSD / sqrt(n())))


# Add site results to output list
  NCRNAnnualWatChem[[site]] = out_data 
  

        
}        
       


#Unlist results and save final dataset
all_water_chem_data = data.frame()
for (i in 1:length(sites)) {
  
  temp = as.data.frame(NCRNAnnualWatChem[[i]]) 
  temp$site = rep(sites[i],nrow(temp))
  all_water_chem_data = rbind(all_water_chem_data,temp)
    
}

# Split the site into three 
split_columns = str_split_fixed(all_water_chem_data$site, "_", 3)
result_df <- data.frame(split_columns)
colnames(result_df) <- c("col1", "Park", "Site")
all_water_chem_data[,9] = result_df$Site
all_water_chem_data$Park = result_df$Park

all_water_chem_data = all_water_chem_data%>%
  mutate(AnnualMedian = round(AnnualMedian,2),
         AnnualMean = round(as.numeric(AnnualMean),2),
         AnnualSD = round(AnnualSD,2),
         Annuallower_ci = round(Annuallower_ci,2),
         Annualupper_ci = round(Annualupper_ci,2))


write.csv(all_water_chem_data,"/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/all_water_chem_data.csv")




```




TEMPORAL CHANGE IN FISH COMMUNITY ABUNDANCE
Uses dataset 'all_fish_count_data.csv' from FishDataCleaning script 
```{r}


##Import data file 

setwd("/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/")
fishData = read.csv("all_fish_count_data.csv")

# Create a dummy variable based on year difference
fishData = fishData %>%
  group_by(site) %>%
  mutate(X = samplingYear - min(samplingYear))

#subset each species
fishData_subset = fishData %>%
  filter(., SubjectTaxonomicName == "Cottus spp.")



#plot the data
ggplot(fishData_subset, aes(x = X, y = TotalCount)) +
  geom_line(size = 1) +
  geom_point(size = 3, shape = 16) +  # Use shape 16 for solid dots
  labs(title = "Fish Species Abundance Over Years",
       x = "Year",
       y = "Number of Individuals") +
  theme_minimal() +
  facet_wrap(~site, scales = "free_y")




```




PLOTTING BOTH CHANGE IN FISH COMMUNITY ABUNDANCE and WATER CHEM TRENDS
Uses dataset 'all_fish_count_data.csv' and 'all_water_chem_data.csv' generated before
```{r}
###Import data file 

setwd("/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/")
Chemdata = read.csv("all_water_chem_data.csv")
Fishdata = read.csv("all_fish_count_data.csv")


# Get unique site names
unique_sites <- unique(Chemdata$site)

# empty lists to store results
filtered_chem_data_list <- list()
filtered_fish_data_list <- list()


for (i in unique_sites) {
  
  #  Chemdata
  filtered_chem_data <- Chemdata %>%
    filter(site == i & AnnualCount > 10 )
  
  #  Fishdata
  filtered_fish_data <- Fishdata %>%
    filter(site == i & TotalCount > 10 )
  
  # store into empty lists
  filtered_chem_data_list[[i]] <- filtered_chem_data
  filtered_fish_data_list[[i]] <- filtered_fish_data
}


PlotChemdata <- bind_rows(filtered_chem_data_list, .id = "site") # Combine results into data frame
PlotFishdata <- bind_rows(filtered_fish_data_list, .id = "site") # Combine results into data frame


######################################SELECT WATER CHEMISTRY OR PARAMATER OF INTEREST#########################################
watChemParam = "Temperature, water"




# Initialize an empty list to store plots
plots_list <- list()

for (i in unique_sites) {
  # Filter Chemdata
  filtered_chem_data <- Chemdata %>%
    filter(site == i & AnnualCount > 10 & CharacteristicName == watChemParam)
  
  # Filter Fishdata
  filtered_fish_data <- Fishdata %>%
    filter(site == i & TotalCount > 10)
  
  # Plot Chemdata
  chem_plot <- ggplot(filtered_chem_data, aes(x = samplingYear, y = AnnualMedian)) +
    geom_point(position = position_dodge(width = 0.3), size = 3) +
    geom_errorbar(aes(ymin = AnnualMedian - Annuallower_ci, ymax = AnnualMedian + Annualupper_ci),
                  position = position_dodge(width = 0.3), width = 0.2) +
    labs(title = paste("Trend in Waterchem - Site:", i),
         x = "Year",
         y = paste(filtered_chem_data[1,"CharacteristicName"], "Annual Median with CI"))
  
  # Plot Fishdata
  fish_plot <- ggplot(filtered_fish_data, aes(x = samplingYear, y = TotalCount, fill = SubjectTaxonomicName)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
    scale_fill_viridis_d() + 
    labs(title = paste("Total Count per Fish Species - Site:", i),
         x = "Year",
         y = "Total Count") +
    theme(legend.position = "top")
  
  # Combine  plots 
  combined_plot <- cowplot::plot_grid(chem_plot, fish_plot, ncol = 1, align = "v", labels = "AUTO")
  

  plots_list[[i]] <- combined_plot
}


# Save the plots to files 
file_path <- "/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Figures/FishChembySite"
for (i in unique_sites) {
  # Save each plot to a file
  ggsave(file.path(file_path, paste("plot_site_", i, watChemParam, ".png", sep = "")),
         plot = plots_list[[i]], device = "png")
}


print(plots_list[[3]])

```

