---
title: "FishCommunityAnalysis_NPS_2023"
author: "MPeipoch"
date: "2023-12-03"
output: pdf_document
---

Necessary packages for the script
```{r, echo=FALSE}
# required packages
req_packs <- c("ggplot2", "dplyr", "vegan", "tidyr","stringr","openxlsx","sf","raster","trend","lubridate")


########### if installed, then install and load ###################################
new_packs <- req_packs[!(req_packs %in% installed.packages()[,"Package"])]
if (length(new_packs) > 0) {
  install.packages(new_packs,dependencies=T)
}

lapply(req_packs, function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    stop(paste("Package", pkg, "could not be loaded. Fix it!"))
  }
  library(pkg, character.only = TRUE)
})




```

ORGANIZING FISH DATA FOR COMMUNITY ANALYSIS:
This section generates two new datasets, one containing general diversity metrics per site and year and another one corresponding to typical species abundance matrix
```{r}

###Import data file 

setwd("/Users/mpeipoch/Library/CloudStorage/GoogleDrive-mpeipoch@stroudcenter.org/My Drive/NPS_FishDataAnalysis")
wdata = read.csv("20231128_wqp_wqx_bss_wq_npsncrn.csv")
  

  
###Create a vector with the unique sites in USNPS NCRN Biological stream survey

sites = unique(wdata$MonitoringLocationIdentifier)



###Create lists of length=sites in which all available fish data and diversity metrics per site is stored in a unique format and file

NCRNFishPopData = list() # Results output--RawData
NCRNFishRichness = list() # Results output--Richness


for (site in sites) {
  
  # Subset data for each site
  temp_data = wdata %>% 
    dplyr::filter(MonitoringLocationIdentifier == site)

    
        temp_data_subset = temp_data %>% #select variable per site and method 
          dplyr::filter(CharacteristicName %in% c("fish - count of individuals captured by species","fish specimens captured electrofishing pass 1","fish specimens captured electrofishing pass 2"))
        #all samplings done post 2018 (by Stroud) are summarized in CharacteristicName=fish specimens captured electrofishing pass 1 or 2; samplings prior to 2018 are summarized by CharacteristicName=fish - count of individuals captured by species, but in both cases the data are expressed as total count of individuals in 'ResultMeasureValue' and id's in 'SubjectTaxonomicName'

 
out_data = temp_data_subset %>%
  mutate(samplingYear =  as.factor(year(temp_data_subset$ActivityStartDate))) %>%
    group_by(samplingYear,SubjectTaxonomicName) %>%
    summarise(TotalCount = sum(as.numeric(ResultMeasureValue),na.rm=T))
  
out_richness = out_data %>%
    group_by(samplingYear) %>%
    summarise(SpeciesRichness = n_distinct(SubjectTaxonomicName),
              ShannonIndex = round(diversity(TotalCount, index = "shannon"),2),
              SimpsonIndex = round(diversity(TotalCount, index = "simpson"),2))


# Add site results to output list
  NCRNFishPopData[[site]] = out_data
  NCRNFishRichness[[site]] = out_richness 
  

        
}        
       


#Unlist results and save final dataset
all_fish_diversity_data = data.frame()
for (i in 1:length(sites)) {
  
  temp = as.data.frame(NCRNFishRichness[[i]]) 
  temp$site = rep(sites[i],nrow(temp))
  all_fish_diversity_data = rbind(all_fish_diversity_data,temp)
    
}

# Split the site into three 
split_columns = str_split_fixed(all_fish_diversity_data$site, "_", 3)
result_df <- data.frame(split_columns)
colnames(result_df) <- c("col1", "Park", "Site")
all_fish_diversity_data[,4] = result_df$Site
all_fish_diversity_data$Park = result_df$Park

write.csv(all_fish_diversity_data,"/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/all_fish_diversity_data.csv")


#Unlist results and save final dataset
all_fish_count_data = data.frame()
for (i in 1:length(sites)) {
  
  temp = as.data.frame(NCRNFishPopData[[i]]) 
  temp$site = rep(sites[i],nrow(temp))
  all_fish_count_data = rbind(all_fish_count_data,temp)
    
}

# Split the site into three 
split_columns = str_split_fixed(all_fish_count_data$site, "_", 3)
result_df <- data.frame(split_columns)
colnames(result_df) <- c("col1", "Park", "Site")
all_fish_count_data[,4] = result_df$Site
all_fish_count_data$Park = result_df$Park

write.csv(all_fish_count_data,"/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/all_fish_count_data.csv")


```


ORGANIZING FISH DATA FOR COMMUNITY ANALYSIS:
This section generates anther diveristy dataset at the Park level to generate the Sankey Plot.
```{r}

###Import data file 

setwd("/Users/mpeipoch/Library/CloudStorage/GoogleDrive-mpeipoch@stroudcenter.org/My Drive/NPS_FishDataAnalysis")
wdata = read.csv("20231128_wqp_wqx_bss_wq_npsncrn.csv")
  

  
###Create a vector with the unique sites in USNPS NCRN Biological stream survey
wdata = wdata %>%
    separate(MonitoringLocationIdentifier, into = c("region", "NCRN_park", "site"), sep = "_")


  
parks = unique(wdata$NCRN_park)



###Create lists of length=sites in which all available fish data and diversity metrics per site is stored in a unique format and file

NCRNFishRichness = list() # Results output--Richness


for (park in parks) {
  
  # Subset data for each park
  temp_data = wdata %>% 
    dplyr::filter(NCRN_park == park)

    
        temp_data_subset = temp_data %>% #select variable per site and method 
          dplyr::filter(CharacteristicName %in% c("fish - count of individuals captured by species","fish specimens captured electrofishing pass 1","fish specimens captured electrofishing pass 2"))
        #all samplings done post 2018 (by Stroud) are summarized in CharacteristicName=fish specimens captured electrofishing pass 1 or 2; samplings prior to 2018 are summarized by CharacteristicName=fish - count of individuals captured by species, but in both cases the data are expressed as total count of individuals in 'ResultMeasureValue' and id's in 'SubjectTaxonomicName'

 
out_data = temp_data_subset %>%
    summarise(SpeciesRichness = n_distinct(SubjectTaxonomicName))


# Add park results to output list
  NCRNFishRichness[[park]] = out_data
  

        
}        
       


#Unlist results and save final dataset
all_fish_diversity_data = data.frame()
for (i in 1:length(parks)) {
  
  temp = as.data.frame(NCRNFishRichness[[i]]) 
  temp$site = rep(parks[i],nrow(temp))
  all_fish_diversity_data = rbind(all_fish_diversity_data,temp)
    
}


write.csv(all_fish_diversity_data,"/Volumes/Stroud Users/Peipoch/Private/dataAnalysis/R/Rstudio/FishAnalysisNCRN/Datasets/all_fish_diversity_data_byPark.csv")


```



















